Detecting ditches using machine learning on high-resolution DEMs

### 2021-01-16 LaTeX ESWA/article.tex
Rewrites up to section 2.4.2 to be more suited for ESWA and more concise.

\title{Detecting ditches using \hl{supervised} learning on high-resolution \hl{digital elevation models}}

@@

\begin{abstract}
{\footnotesize
\hl{Drained wetlands can constitute a large source of greenhouse gas emissions, but the drainage networks in these wetlands are largely unmapped, and better maps are needed to aid in forest management decisions, and to predict the climate consequences.} We develop a method for detecting ditches in high resolution digital elevation models derived from LiDAR scans. \hl{Thresholding methods using digital terrain indices can be used to detect ditches,} however, a single threshold generally does not capture the variability in the landscape, and generates many false positives and negatives. We hypothesise that, by combining the digital terrain indices using \hl{supervised} learning, we can improve ditch detection at a landscape-scale. In addition to digital terrain indices, additional features are generated by transforming the data to include neighbouring cells \hl{for better ditch predictions}. \hl{A Random Forests classifier} is used to locate the ditches, and \hl{its probability output} is processed to remove noise, and binarised to produce the final ditch prediction. The confidence interval for the Cohen's Kappa index ranges \textbf{[0.655 , 0.781]} between the evaluation plots with a confidence level of \textbf{95\%}. The study demonstrates that combining information from  a suite of digital terrain indices using machine learning provides an effective technique for automatic ditch detection at a landscape-scale\hl{, aiding in both practical forest management and in combatting climate change}.}
\end{abstract}

@@

\subsection{Study area and dataset}
The quaternary deposits in the Krycklan Study Catchment are dominated by till (51\%) and sorted sediments (30\%). The catchment ranges in elevation from 114 to 405 m above sea level. Forest covers 87\% of the catchment, mires 9\%, thin soils 7\%, and rock outcrops 1\%. The land use is dominated by forestry; approximately 25\% of the Krycklan catchment has been protected since 1922, but  the rest of the area is mostly second growth forest \citep{krycklancatchment}.

A historical survey of the catchment (Gudrun Norstedt, unpublished) recorded  that the ditches in the catchment were dug between 1900 and 1993. Initially, the ditches were dug to drain the mires for hay production. Later on, wet areas were drained to attempt to turn them into productive forest lands \citep{paivanen}. However, research later showed that many of the drained mires were not productive for forestry due to lack of nutrients \citep{sikstrom}.

The LiDAR dataset used in this study was produced by TerraTec Sweden AB in August 2015, on demand of the Department of Forest Resource Management at the Swedish University of Agricultural Sciences, and is freely available online \citep{dataset}. The average density of the dataset was 20 pulses per $m^2$. Ground point classification was performed with lasground (included in LAStools, a software suite produced by rapidlasso GmbH). The DEM was created with blast2dem (included in LAStools), with a cell size of $0.5*0.5$ m. Only ground points were imported to generate a DEM of the ground.

\subsection{Digitising the ground truth}
To train and validate the machine learning models, a vector layer of all ditches in the Krycklan Catchment \citep{krycklancatchment} was digitised. The study catchment comprises several types of ditches including road ditches, forest ditches, and agricultural ditches. A multidirectional hillshade of the 0.5 m DEM was calculated in ArcGIS Pro to locate the ditches. From this layer, most of the ditches could be easily identified. We conducted the digitisation at a scale of 1:600 to ensure all the digitised ditches properly align the actual ditch on the ground and do not coincide with the edge of the banks. In addition, both current and historical records and orthophotos dating back to the 1960's were consulted to help locate all the ditches. In uncertain areas, the ditches were verified in the field. The stream network in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan} shows a network with a 8 hectare flow initiation threshold. The digitisation of the ditch networks showed that there are 107 km of road ditches and 208 km of forest/agricultural ditches in the catchment (\hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}). The stream network in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}, which represents median flow conditions, was 197 km.

\begin{figure}[!htb]
    \centering
    \includegraphics[width=1\linewidth]{./images/Krycklan_lo.jpg}
    \caption{Map of Krycklan showing ditches and stream channels, as well as our division into 21 subsections. The 10 subsections with a white border were used for developing features and optimising parameters before the experiment. The 11 subsections with a black border were used in the experiment for evaluation with 11-fold cross validation. Each rectangular subsection represents $2997 \cdot 2620$ pixels (roughly 196 hectare).}
    \label{fig:swedenkrycklan}
\end{figure}

From the ditch mapping, the vector layer was rasterised so that it could be compared to the automatically derived ditches from the $0.5*0.5$ m DEM. Our field observations have shown that the  width of most ditches within the catchment ranges between 0.5 and 3.5 m. To ensure that the model received all ditch pixels correctly in the training phase, we widened the ditches so that all pixels within a radius of three pixels (1.5 metres) of the vectors were labelled as ditch pixels. \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{a} shows the ditches rasterised from vectors and \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{b} shows the ditches after widening. The data in \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{b} is the labelled data that was used to train the machine learning models. The ditches all have a generalised width based on a perceived average of ditch width in the area. Due to all ditches varying in width, it was impossible to produce a perfect representation of each ditch, but this widening helped ensure that almost all ditch pixels were covered by the labels.

To match the labels with out predictions (where we downscale the predictions into grids to remove outliers), the ground-truth was also downscaled into grids of $6*6$ pixels (9 $m^2$) each. Each grid cell that contained at least 25\% ditch pixels was labelled as a ditch, see \hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{c}. This layer was used to ground-truth the machine learning models.

\begin{figure} [!htb]
    \centering
    a{
        \resizebox*{3.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_A_lo.jpg}}}\hspace{5pt}
    b{
        \resizebox*{3.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_B_lo.jpg}}}
    c{
        \resizebox*{3.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_C_lo.jpg}}}
    \caption{Processing of ditch labels. \textbf{a: }Rasterised ditches with a width of one pixel. \textbf{b: }Ditches after a widening process, seven pixels (3.5 metres) wide. Used as input when training the models. \textbf{c: }Ditches after a grid zone conversion. Used for evaluating the results from the ditch detector.} \label{sample-figure}
    \label{fig:ditchpreprocess}
\end{figure}

\subsection{Extracting ditches with digital terrain indices}
From the DEM, several digital terrain indices were calculated. The strengths and weaknesses of these indices have been obtained from observing their performance in different topological environments:
\begin{itemize} \label{data_attributes}

  \item \textbf{Sky View Factor} \newline
    The Sky View Factor represents how much of the sky that is visible from a certain point on the ground \citep{zaksek}. A point in a ditch would yield a low value, as more of the sky is obscured by the ditch bank. However, it tends to produce false positives in steep terrain. The Sky View Factor was calculated in SAGA GIS, using a search radius of 10 metres to ensure that it was large enough to capture the ditches, but small enough to not include large scale topographical features such as hills.

  \item \textbf{Impoundment Index}\newline
    The Impoundment Index in Whitebox Tools \citep{whiteboxtools} can be used as a measure of flow constriction, which is useful for mapping channels, such as ditches. When tuned to output the flooded volume, it effectively maps the area and average depth of the reservoir that would be created by placing a dam of a user-specified length at each grid cell in the DEM. Here, we used dam height calculated with a dam length of 3 metres to cover the entire ditches. Typical false positives with this method are natural stream channels with a clear incised channel.

@@

    \item \textbf{Slope} \newline
    The Slope index represents the degree of slope at a certain point, with a value ranging from $0^{\circ}$ to $90^{\circ}$. It was calculated on the 0.5 m grid, using ArcGIS \citep{EsriArcGisBook}. This index contains no information about the direction of the slope.
\end{itemize}

To benchmark our method against a non-machine learning method, we binarised the original four digital terrain indices using thresholds. Different threshold values were tested and the value that yielded the highest Cohen's Kappa score for each respective index was chosen. This test was conducted for the 11 subsections described in \ref{trainingvalidationdatasets}. The Sky View Factor index was classified to only include values below 0.955, the Impoundment Index Dam Height using values above 0.27 $m$, the HPMF index using values below -0.18 m, and the Slope index using values above $14 ^{\circ}$ (Table 2).

@@

\subsubsection{Processing the digital terrain indices}
We examined how different  features affected the prediction of the machine learning models. Several possible data manipulation methods could theoretically produce a better prediction. To improve the prediction, steps were taken where neighbouring pixels were included to give a representation of the area surrounding a specific pixel, similar to the approach of \citet{roelens}.

The digital terrain indices provided a satisfactory foundation for the models, but lacked in the generalisability of their predictions. More diverse features were extracted using simple statistical aggregates such as mean, median, min, max, and standard deviation. Using these statistical aggregations with the help of neighbouring areas around pixels aided in pruning pixels with outlier values, often smoothing out the data to represent ditches more accurately on a per-pixel basis. These variables were calculated by gathering all data points in different circular radii around the studied pixel, before computing one of the statistical aggregations (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{b}, \hyperref[fig:features]{c}, \hyperref[fig:features]{f}, and \hyperref[fig:features]{g}).

In addition to the general features, a number of custom features were also developed.  Some of these custom variables were produced by statistical aggregation of multiple indices, whereas in some cases, they were generated by using thresholds to exclude, lower, or amplify the values of pixels that indicated ditches or non-ditches (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{d} and \hyperref[fig:features]{h}).

Both HPMF and Sky View Factor were used with the image processing \textit{gabor filter}, which can be used to detect lines of a certain orientation in an image \citep{gabor}. 30 Gabor filters which were rotated in different angles and with different frequencies were combined to detect lines, amplifying ditches by utilising the fact that ditches have a linear elongated shape (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{e}).

\label{impoundmentstreamremoval}
The raw Impoundment Index dam height was used to create a mask, attempting to mask out streams, but retain ditches. Only areas with a relatively high value, which would indicate that these areas most likely contained streams, were masked out. After widening the resulting area, this mask was used to remove streams from all the aforementioned custom features, generating one new feature from each. The features marked with \textit{streams removed} in \hyperref[featuretable]{Table} \ref{featuretable} make use of this filter (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{i}).

Overall, 81 features were extracted from the terrain indices. In a pilot study, we compared several different algorithms (Random Forests, Extreme Gradient Boosting, Naive Bayes, and Support Vector Machines), and it was found that Random Forests produced the most accurate results, here we therefore only present the results from the Random Forest model. We then conducted a sub-experiment to find the best features, as well as the optimal number of variables to use. The variable importance score in Random Forests was used to select the most important features and it was found that using the top 40 features produced the best results.

@@

\subsubsection{Training and validation datasets}
\label{trainingvalidationdatasets}
To develop and evaluate our ditch detector, the raster and ditch label data of Krycklan were manually divided into 21 subsections. \hl{This was necessary because our method relies on the area surrounding a pixel being examined, meaning it is not possible to select random pixels for training and testing.} Each subsection represents an area of roughly 196 hectare. From this division, 11 of the subsections were put aside as hold-out data, only for use in the experiment to evaluate the performance of the predictions. The remaining 10 subsections were used solely before the experiment to develop the ditch detector to test different ways of preparing the data. This allowed the ditch detector to be evaluated on  independent data and  strengthened the validity of the experiment. \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan} shows which subsections were used for development and evaluation respectively. \hl{The 11 hold-out subsections were used as folds in an 11-fold cross validation in the final experiment.}

@@

      Slope mean & 6 \\
      Slope median & 6 \\
      Slope standard deviation & 4, 6 \\
      Slope min & 2, 4, 6 \\
      Slope non-ditch amplification & - \\
      \hline
    \end{tabular}}
    \caption{ The 40 features used when training the models.
    \newline \textsuperscript{a} Displays algorithm used to produce the feature. \newline
        \textsuperscript{b} Represents the radius of the circular mask (if one was used) to determine \newline what neighbouring pixels to use in an algorithm. The radii represents pixels \newline \mbox{with a 0.5 m resolution.\itshape\ignorespaces}}
    \label{featuretable}
\end{table}

@@

\begin{table}[!htb]
\centering
    {\begin{tabular}{l|cccc}
        \textbf{Metric} & \textbf{Sky View Factor} & \textbf{Impoundment} & \textbf{HPMF} & \textbf{Slope}\\
        \hline
        Accuracy \%     & 83.67 & 97.07 & 97.45 & 82.66 \\
        Recall \%       & 44.54 & 29.98 & 25.31 & 34.83 \\
        Precision \%    &{ 4.00}  & 18.51 & 20.10 & { 2.99} \\
        $\kappa$ rating & 0.048 & 0.215 & 0.211 & 0.029 \\
        AUPRC & 0.247 & 0.248 & 0.232 & 0.194 \\
        \hline
    \end{tabular}}
    \caption{Metrics from the total results of the four digital terrain indices experiments\hl{, where only a sole terrain index is used with a threshold to detect ditches}.}
    \label{recreatedpredictionperformance}
\end{table}

@@

Although there is room for further improvement, the methodology is a significant step forward as it performed on a par with previous studies conducted in more homogeneous agricultural landscapes. Our approach enables accurate ditch detection in complex forested landscapes with tree canopy and many natural streams. Future work could introduce more features and post-processing steps, such as using pathfinding algorithms to fill gaps in the ditch model or shape indices to remove stream channels from the prediction. Moving towards image segmentation instead of pixel classification of tabular pixel values could also be a suitable future approach. Our developed methodology can be adjusted to apply to any setting for artificial ditch detection, reducing time and labour investment compared to the conventional manual mapping techniques. \hl{The accurate ditch maps can both significantly contribute to practical forestry and operational land management, as well as allow stakeholders including government agencies, forest companies, and regional and local authorities to predict the climate consequences of potential management actions, which will be critical in reaching global sustainability goals.}

\section*{CRediT authorship contribution statement}
\hlc{Jag tolkar det som att denna sektion är frivillig. De flesta artiklar jag har hittat listar här vad alla enskilda författare har bidragit med. Kategorier som andra har med här är: Conceptualisation, Data curation, Formal analysis, Investigation, Methodology, Software, Validation, Visualisation, Writing - original draft, Writing - review & editing, Supervision, Project administration}

@@

\section*{Acknowledgements}
We wish to thank Dr. Siddhartho Paul at the Swedish University of Agricultural Sciences, whose suggestions helped improve and clarify this manuscript. We also thank Sigurd Israelsson at the School of Engineering at J\"onk\"oping University for lending hardware resources for the experiments in this study. This work was supported by VINNOVA, EU InterReg Baltic Sea project WAMBAF tools and Formas. \hlc{ \textbf{ Här ska varje grant definieras med "grant number[xxxx] efter varje aktör"} }

>>>>

\title{Detecting ditches using supervised learning on high-resolution digital elevation models}

@@

\begin{abstract}
{\footnotesize
Drained wetlands can constitute a large source of greenhouse gas emissions, but the drainage networks in these wetlands are largely unmapped, and better maps are needed to aid in forest management decisions, and to predict the climate consequences. We develop a method for detecting ditches in high resolution digital elevation models derived from LiDAR scans. Thresholding methods using digital terrain indices can be used to detect ditches, however, a single threshold generally does not capture the variability in the landscape, and generates many false positives and negatives. We hypothesise that, by combining the digital terrain indices using supervised learning, we can improve ditch detection at a landscape-scale. In addition to digital terrain indices, additional features are generated by transforming the data to include neighbouring cells for better ditch predictions. A Random Forests classifier is used to locate the ditches, and its probability output is processed to remove noise, and binarised to produce the final ditch prediction. The confidence interval for the Cohen's Kappa index ranges \textbf{[0.655 , 0.781]} between the evaluation plots with a confidence level of \textbf{95\%}. The study demonstrates that combining information from  a suite of digital terrain indices using machine learning provides an effective technique for automatic ditch detection at a landscape-scale, aiding in both practical forest management and in combatting climate change.}
\end{abstract}

@@

\subsection{Study area and dataset}
The quaternary deposits in the Krycklan Study Catchment are dominated by till (51\%) and sorted sediments (30\%). The catchment ranges in elevation from 114 to 405 m above sea level. Forest covers 87\% of the catchment, mires 9\%, thin soils 7\%, and rock outcrops 1\%. The land use is dominated by forestry; approximately 25\% of the Krycklan catchment has been protected since 1922, but  the rest of the area is mostly second growth forest \citep{krycklancatchment}.  A historical survey of the catchment (Gudrun Norstedt, unpublished) recorded  that the ditches in the catchment were dug between 1900 and 1993. Initially, the ditches were dug to drain the mires for hay production. Later on, wet areas were drained to attempt to turn them into productive forest lands \citep{paivanen}. However, research later showed that many of the drained mires were not productive for forestry due to lack of nutrients \citep{sikstrom}.

The LiDAR dataset used in this study was produced by TerraTec Sweden AB in August 2015, on demand of the Department of Forest Resource Management at the Swedish University of Agricultural Sciences, and is freely available online \citep{dataset}. The average density of the dataset is 20 pulses per $m^2$. Ground point classification was performed with lasground (included in LAStools, a software suite produced by rapidlasso GmbH). The DEM was created with blast2dem (included in LAStools), with a cell size of $0.5*0.5$ m. Only ground points were imported to generate a DEM of the ground.

\subsection{Digitising the ground truth}
For training and validation, a vector layer of all ditches in the Krycklan Catchment \citep{krycklancatchment} was digitised. The study catchment comprises several types of ditches including road ditches, forest ditches, and agricultural ditches. A multidirectional hillshade of the 0.5 m DEM was calculated in ArcGIS Pro, and from this layer, most of the ditches could be easily identified. We conducted the digitisation at a scale of 1:600 to ensure that all the digitised ditches are properly aligned with the actual ditch on the ground, and do not coincide with the edge of the banks. In addition, both current and historical records and orthophotos dating back to the 1960's were consulted to help locate all the ditches. In uncertain areas, the ditches were verified in the field. The digitisation of the ditch networks showed that there are 107 km of road ditches and 208 km of forest/agricultural ditches in the catchment (\hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan}). The stream network in \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan} shows a 197 km long network with a 8 hectare flow initiation threshold, representing median flow conditions.

\begin{figure}[!htb]
    \centering
    \includegraphics[width=1\linewidth]{./images/Krycklan_lo.jpg}
    \caption{Map of Krycklan showing ditches and stream channels, as well as our division into 21 subsections of $2997 \cdot 2620$ pixels (roughly 196 hectare). The 10 subsections with a white border were used for developing features and optimising parameters before the experiment. The 11 subsections with a black border were used in the experiment for evaluation with 11-fold cross validation.}
    \label{fig:swedenkrycklan}
\end{figure}

From the ditch mapping, the vector layer was rasterised with a resolution of $0.5*0.5$ metres for use as a ground-truth for our ditch detector. Our field observations have shown that most of the ditches in the catchment range between 0.5 and 3.5 metres in width. To ensure that all ditch pixels were labelled correctly for the model in the training phase, we labelled all pixels within three pixels (1.5 metres) of the vectors as ditch pixels, making the labels 3.5 metres wide (\hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess}). Due to all ditches varying in width, it was impossible to produce a perfect representation of each ditch, but this widening helped ensure that almost all ditch pixels were covered by the labels. To match the labels with our predictions (where we downscale the predictions into grids to remove outliers), the ground-truth was also downscaled into grids of $6*6$ pixels (9 $m^2$) each by labelling each grid cell containing at least 25\% ditch pixels as a ditch (\hyperref[fig:ditchpreprocess]{Figure} \ref{fig:ditchpreprocess} \hyperref[fig:ditchpreprocess]{c}).

\begin{figure} [!htb]
    \centering
    a{
        \resizebox*{3.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_A_lo.jpg}}}\hspace{5pt}
    b{
        \resizebox*{3.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_B_lo.jpg}}}
    c{
        \resizebox*{3.5cm}{!}{\includegraphics{./images/publ_ditch_preprocess_C_lo.jpg}}}
    \caption{Processing of ditch labels. \textbf{a: }Rasterised ditches with a width of one pixel. \textbf{b: }Ditches after a widening process, seven pixels (3.5 metres) wide. Used as labels when training the models. \textbf{c: }Ditches after a grid zone conversion. Used for evaluating the results from the ditch detector.} \label{sample-figure}
    \label{fig:ditchpreprocess}
\end{figure}

\subsection{Extracting ditches with digital terrain indices}
From the DEM, several digital terrain indices were calculated:
\begin{itemize} \label{data_attributes}

  \item \textbf{Sky View Factor} \newline
    The Sky View Factor represents how much of the sky that is visible from a certain point on the ground \citep{zaksek}. A point inside a ditch would produce a low value, as more of the sky is obscured by the ditch bank. Sky View Factor tends to produce false positives in steep terrain. The index was calculated in SAGA GIS, using a search radius of 10 metres to ensure that it was large enough to capture the ditches, but small enough to not include large scale topographical features such as hills.

  \item \textbf{Impoundment Index}\newline
    The Impoundment Index in Whitebox Tools \citep{whiteboxtools} can be used as a measure of flow constriction, which is useful for mapping channels such as ditches. When tuned to output the flooded volume, it effectively maps the area and average depth of the reservoir that would be created by placing a dam of a user-specified length at each grid cell in the DEM. Here, we used dam height calculated with a dam length of 3 metres to attempt to cover the entire ditches. Typical false positives with this index are natural stream channels with a clear incised channel.

@@

    \item \textbf{Slope} \newline
    The Slope index represents the degree of slope at a certain point, with a value ranging from $0^{\circ}$ to $90^{\circ}$ with no information about the direction of the slope. It was calculated on the 0.5 m grid, using ArcGIS \citep{EsriArcGisBook}.
\end{itemize}

To benchmark our method against a non-machine learning method, we binarised the original four digital terrain indices using thresholds. Different threshold values were tested and the value that yielded the highest Cohen's Kappa score for each respective index was chosen. This test was conducted for the 11 subsections described in \ref{trainingvalidationdatasets}. The Sky View Factor index was classified to only include values below 0.955, the Impoundment Index Dam Height using values above 0.27 $m$, the HPMF index using values below -0.18 m, and the Slope index using values above $14 ^{\circ}$ (\autoref{recreatedpredictionperformance}).

@@

\subsubsection{Feature engineering}

The digital terrain indices provided a satisfactory foundation for the models, but could not always generalise well to outlier situations. To give a representation of the area surrounding a specific pixel, we extracted more diverse features using simple statistical aggregates such as mean, median, min, max, and standard deviation in different circular radii around the pixel (similar to the approach of \citet{roelens}). This aided in pruning pixels with outlier values, often smoothing out the data to represent ditches more accurately on a per-pixel basis (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{b}, \hyperref[fig:features]{c}, \hyperref[fig:features]{f}, and \hyperref[fig:features]{g}). Additionally, several features were generated by using thresholds on one or multiple indices to exclude, lower, or amplify the values of pixels that indicated ditches or non-ditches (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{d} and \hyperref[fig:features]{h}).

Features were generated for both HPMF and Sky View Factor using the image processing \textit{gabor filter}, which can be used to detect lines of a certain orientation in an image \citep{gabor}. 30 Gabor filters which were rotated in different angles and with different frequencies were combined to detect lines, amplifying ditches by utilising the fact that ditches have a linear elongated shape (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{e}).

\label{impoundmentstreamremoval}
In an attempt to exclude streams from the predictions, the Impoundment Index was used to mask out areas with a very large dam height (which would indicate that the area most likely contained a stream). This mask was applied on several features to smooth out the values under the mask, generating one new feature from each of them. The features marked with \textit{streams removed} in \hyperref[featuretable]{Table} \ref{featuretable} make use of this filter (\hyperref[fig:features]{Figure} \ref{fig:features}: \hyperref[fig:features]{i}).

Overall, 81 features were extracted from the terrain indices. In a pilot study, we compared several different algorithms (Random Forests, Extreme Gradient Boosting, Naive Bayes, and Support Vector Machines), and it was found that Random Forests produced the most accurate results. We then conducted a sub-experiment using Random Forests' feature importance to find the best features, as well as the optimal number of features to use, resulting in a final feature set of 40 features (\autoref{featuretable}).

@@

\subsubsection{Training and validation datasets}
\label{trainingvalidationdatasets}
To develop and evaluate our ditch detector, the raster and ditch label data of Krycklan were manually divided into 21 subsections. This was necessary because our method relies on the area surrounding a pixel being examined, meaning it is not possible to select random pixels for training and testing. Each subsection represents an area of roughly 196 hectare. From this division, 11 of the subsections were put aside as hold-out data, only for use in the experiment to evaluate the performance of the predictions. The remaining 10 subsections were used solely before the experiment to develop the ditch detector to test different ways of preparing the data. This allowed the ditch detector to be evaluated on  independent data and  strengthened the validity of the experiment. \hyperref[fig:swedenkrycklan]{Figure} \ref{fig:swedenkrycklan} shows which subsections were used for development and evaluation respectively. The 11 hold-out subsections were used as folds in an 11-fold cross validation in the final experiment.

@@

      Slope mean & 6 \\
      Slope median & 6 \\
      Slope standard deviation & 4, 6 \\
      Slope min & 2, 4, 6 \\
      Slope non-ditch amplification & - \\
      \hline
    \end{tabular}}
    \caption{ The 40 features used to train the models.
    \newline \textsuperscript{a} The algorithm used to produce the feature. \newline
        \textsuperscript{b} Represents the radius of the circular mask (if one was used) to determine which neighbouring pixels to use in the aggregation. The radii represents pixels with a 0.5 m resolution.}
    \label{featuretable}
\end{table}

@@

\begin{table}[!htb]
\centering
    {\begin{tabular}{l|cccc}
        \textbf{Metric} & \textbf{Sky View Factor} & \textbf{Impoundment} & \textbf{HPMF} & \textbf{Slope}\\
        \hline
        Accuracy \%     & 83.67 & 97.07 & 97.45 & 82.66 \\
        Recall \%       & 44.54 & 29.98 & 25.31 & 34.83 \\
        Precision \%    &{ 4.00}  & 18.51 & 20.10 & { 2.99} \\
        $\kappa$ rating & 0.048 & 0.215 & 0.211 & 0.029 \\
        AUPRC & 0.247 & 0.248 & 0.232 & 0.194 \\
        \hline
    \end{tabular}}
    \caption{Metrics from the total results of the four digital terrain indices experiments, where only a sole terrain index is used with a threshold to detect ditches.}
    \label{recreatedpredictionperformance}
\end{table}

@@

Although there is room for further improvement, the methodology is a significant step forward as it performed on a par with previous studies conducted in more homogeneous agricultural landscapes. Our approach enables accurate ditch detection in complex forested landscapes with tree canopy and many natural streams. Future work could introduce more features and post-processing steps, such as using pathfinding algorithms to fill gaps in the ditch model or shape indices to remove stream channels from the prediction. Moving towards image segmentation instead of pixel classification of tabular pixel values could also be a suitable future approach. Our developed methodology can be adjusted to apply to any setting for artificial ditch detection, reducing time and labour investment compared to the conventional manual mapping techniques. The accurate ditch maps can both significantly contribute to practical forestry and operational land management, as well as allow stakeholders including government agencies, forest companies, and regional and local authorities to predict the climate consequences of potential management actions, which will be critical in reaching global sustainability goals.

\section*{CRediT authorship contribution statement}
\textbf{Jag tolkar det som att denna sektion är frivillig. De flesta artiklar jag har hittat listar här vad alla enskilda författare har bidragit med. Kategorier som andra har med här är: Conceptualisation, Data curation, Formal analysis, Investigation, Methodology, Software, Validation, Visualisation, Writing - original draft, Writing - review & editing, Supervision, Project administration}

@@

\section*{Acknowledgements}
We wish to thank Dr. Siddhartho Paul at the Swedish University of Agricultural Sciences, whose suggestions helped improve and clarify this manuscript. We also thank Sigurd Israelsson at the School of Engineering at J\"onk\"oping University for lending hardware resources for the experiments in this study. This work was supported by VINNOVA, EU InterReg Baltic Sea project WAMBAF tools and Formas. \textbf{ Här ska varje grant definieras med "grant number[xxxx] efter varje aktör"}

### 2021-01-16 LaTeX/test.tex
Rewrites up to section 2.4.2 to be more suited for ESWA and more concise.

\author{Anonymised.}

@@

%Detection of artificial ditches in a study area with substantial natural stream coverage is challenging but highly desired for effective forest management. This study developed and applied a methodology using multiple terrain indices and machine learning for automatic ditch detection at the Krycklan catchment in Sweden. Despite a substantial vegetation (87\% forest cover), our method achieved good results; the confidence interval for the Cohen's Kappa index ranged [0.655 , 0.781] between the evaluation plots with a confidence level of 95\%. With the developed input variables and extensive post-processing, we managed to remove many stream channels and fill gaps in the ditches where the LiDAR scans were weak. However, these issues are still present to some extent. Future work could introduce more input variables and post-processing steps, such as using pathfinding algorithms to fill gaps in the ditch model, or shape indices to remove stream channels from the prediction. Moving towards image segmentation with grids of pixels instead of pixel classification of tabular pixel values could also be a suitable future approach. With minor adjustments, our proposed method can be employed in a general setting, reducing manual labour and cost, and enabling faster and easier mapping of ditches.

@@

\label{lidartodem}
The LiDAR dataset used in this study was produced by (anonymised), and is freely available online at \href{https://gis.krycklan.se/}{https://gis.krycklan.se/}. The average density of the dataset was 20 pulses per $m^2$.  Ground point classification was performed with lasground (included in LAStools, a software suite produced by rapidlasso GmbH). The DEM was created with blast2dem (included in LAStools),  with cell size of $0.5*0.5$ m. Only ground points were imported  to generate a DEM of the ground. This was conducted for 488 las-files of the study area that were mosaicked to one large file using Mosaic to new raster in Arc GIS Pro \citep{EsriArcGisBook}.

\section*{Acknowledgements}
Anonymised.

>>>>

\author{
\name{Filip Andersson\textsuperscript{a,1}, Jonatan Flyckt\textsuperscript{a,1}, Niklas Lavesson\textsuperscript{a}\thanks{CONTACT Niklas Lavesson. Email: niklas.lavesson@ju.se}, Liselott Nilsson\textsuperscript{b}, and Anneli M. \AA gren\textsuperscript{c}} \affil{\textsuperscript{a}J\"onk\"oping Artificial Intelligence Laboratory, J\"onk\"oping University, Gjuterigatan 5, 551 11, J\"onk\"oping, Sweden; \textsuperscript{b}Forest Department, Swedish Forest Agency, Skeppargatan 17, 931 32 Skellefte\aa, Sweden; \textsuperscript{c}Department of Forest Ecology and Management, Swedish University of Agricultural Sciences, SLU, Skogsmarksgr\"and 17, 901 83, Ume\aa, Sweden}
\footnote{\textsuperscript{1} Filip Andersson and Jonatan Flyckt contributed equally to this work as first authors.}}

@@

@@

\label{lidartodem}
The LiDAR dataset used in this study was produced by TerraTec Sweden AB in August 2015, on demand of the Department of Forest Resource Management at the Swedish University of Agricultural Sciences, and is freely available online at \href{https://gis.krycklan.se/}{https://gis.krycklan.se/}. The average density of the dataset was 20 pulses per $m^2$.  Ground point classification was performed with lasground (included in LAStools, a software suite produced by rapidlasso GmbH). The DEM was created with blast2dem (included in LAStools),  with cell size of $0.5*0.5$ m. Only ground points were imported  to generate a DEM of the ground. This was conducted for 488 las-files of the study area that were mosaicked to one large file using Mosaic to new raster in Arc GIS Pro \citep{EsriArcGisBook}.

\section*{Acknowledgements}
We wish to thank the two anonymous reviewers and Dr. Siddhartho Paul at the Swedish University of Agricultural Sciences, whose suggestions helped improve and clarify this manuscript. We also thank Sigurd Israelsson at the School of Engineering at J\"onk\"oping University for lending hardware resources for the experiments in this study. The work has been funded by VINNOVA, EU InterReg Baltic Sea project WAMBAF tools and Formas.
